### 2016-06-22 Projects/DelCo_2012/Report/DelCo_2012.tex
updates to CAFS, added pdfs that are one off plots

\headerC{
\flushright{
Craig G. Weinschenk \\
Daniel Madrzykowski \\
Keith M. Stakes \\
Joseph M. Willi \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\vspace*{2\baselineskip}
February 2016}}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (February 2016) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAF for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAFS for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAFS and plain water to evaluate the effectiveness of CAFS for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAFS and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2012,LaPolla:2012}. The research findings are provide information with the potential to improve the safe and effective use of CAFS for structural firefighting.

@@

The 88 gas cooling experiments were completed as part of 9 separate tests in the burn building. During each test, the fire was ignited and then developed until the temperature 1.83~m (6~ft) below the ceiling in the gas cooling room reached or exceeded 250~$^{\circ}$C (482~$^{\circ}$F). Once this criteria was met, gas cooling began.

\begin{figure}[ht!]
	\includegraphics[width=\columnwidth]{../Figures/Gas_Cooling/GCSeries4_TC_A3}
	\caption{Time series of several gas cooling events in a continuous experiment}
	\label{fig:gas_cooling_exp4}
\end{figure}

To compare the effectiveness of the suppression agents (CAFS, water), the temperatures from the thermocouple tree which was not directly impacted by the streams was analyzed. Thermocouples that were greater than 2.13~m (7~ft) below the ceiling (the bottom 3 thermocouples) were not included because of the minimal temperature readings throughout the tests. Gas cooling was determined by taking the normalized maximum change in temperature ($\hat{T}$) over the test interval ($T_{max}$ is peak temperature before suppression and $T_{min}$ is minimum temperature after suppression):
\begin{equation}
\label{eq:norm_temp}
\hat{T} = \frac{T_{max}-T_{min}}{T_{max}}
\end{equation}
The normalized temperature difference is taken because of the variation in pre-suppression temperature within replicates of a particular suppression agent and between suppression agents.

For the 7/8~in solid stream tests there were 7 replicate conducted for both water and CAFs suppression. The normalized temperature differences for the 8 thermocouples analyzed from the water experiments were compared against the same normalized temperature differences from the CAFs tests. These data points are visualized through a scatterplot, Fig.~\ref{fig:CAFS_Water_Comp_7_8}.

@@

The circles in Fig.~\ref{fig:CAFS_Water_Comp_7_8} are the normalized temperature measure points. If water and CAFs had the exact same normalized temperature differences, all of the points would fall along solid black line in the figure. Circles that vary from that line represent measurements where either water or CAFS cooled more signifcantly than the other agent. The gray area surrounding the solid black line, represents measurment uncertainty (\ref{subsec:Uncertainty}). For data that fall within this area, there is an inability to determine which agent was more or less effective than the other. The red dashed line is the line the fits the data representing the minimal difference between the two suppression agents by a linear approximation of the data. Since this line falls within the gray area there is an inability to distinguish, with statistical confidence, a difference between water or CAFs to cool gases from a 7/8~in solid stream.

@@

>>>>

\headerB{
Examination of Compressed Air Foam (CAF) for Interior Firefighting
}

\headerC{
\flushright{
Craig G. Weinschenk $^+$ \\
Daniel Madrzykowski $^*$ \\
Keith M. Stakes $^*$ \\
Joseph M. Willi \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\vspace*{2\baselineskip}
July 2016}

\vspace*{2\baselineskip}
\textit{$^+$\small Currently with JENSEN HUGHES}
\\
\textit{$^*$\small Currently with UL FSRI}
}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (July 2016) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAF for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAFS for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAFS and plain water to evaluate the effectiveness of CAFS for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAFS and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. The research findings are provide information with the potential to improve the safe and effective use of CAFS for structural firefighting.

@@

The 88 gas cooling events were completed as part of 9 separate tests in the burn building. During each test, the fire was ignited and then developed until the temperature 1.83~m (6~ft) below the ceiling in the gas cooling room reached or exceeded 250~$^{\circ}$C (482~$^{\circ}$F). Once this criteria was met, gas cooling began. Figure~\ref{fig:gas_cooling_exp4} shows 19 cooling events at one thermocouple array during a continuous test.

\begin{figure}[ht!]
	\includegraphics[width=\columnwidth]{../Figures/Gas_Cooling/GCSeries4_TC_A3}
	\caption{Time series of several gas cooling events in a continuous test}
	\label{fig:gas_cooling_exp4}
\end{figure}

There were three thermocouple arrays in the gas cooling room (adjacent to the burn room in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions}). Figures~\ref{fig:gas_cooling_sub1} and \ref{fig:gas_cooling_sub5} show the variation of the impact of the suppression stream on cooling location for three consecutive water and CAFS tests, respectively. For both sets a solid 7/8 in nozzle in the middle position was used. The data shows that thermocouples in the plot in the upper right of both figures has more significant temperature drops than the other two arrays; this is likely a result of that thermocouple array being directly ``hit'' by the suppression stream.

\begin{figure}[ht!]
	\includegraphics[width=.45\columnwidth]{../Figures/Gas_Cooling/GCSeries12_TC_A1}
	\includegraphics[width=.45\columnwidth]{../Figures/Gas_Cooling/GCSeries12_TC_A2}
	\includegraphics[width=.45\columnwidth]{../Figures/Gas_Cooling/GCSeries12_TC_A3}
	\caption{Time series of several gas cooling events in a continuous test}
	\label{fig:gas_cooling_sub1}
\end{figure}

\begin{figure}[ht!]
	\includegraphics[width=.45\columnwidth]{../Figures/Gas_Cooling/GCSeries52_TC_A1}
	\includegraphics[width=.45\columnwidth]{../Figures/Gas_Cooling/GCSeries52_TC_A2}
	\includegraphics[width=.45\columnwidth]{../Figures/Gas_Cooling/GCSeries52_TC_A3}
	\caption{Time series of several gas cooling events in a continuous test}
	\label{fig:gas_cooling_sub5}
\end{figure}

To better quantify the effectiveness of the suppression agents (CAFS, water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for the 30 replicate tests, 15 water and 15 CAFS, for all streams aimed at all positions.

\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Gas_Cooling/Combined_all_scatter}
	\caption{CAFS vs Water Gas Cooling Comparison for all replicate tests}
	\label{fig:combined_all}
\end{figure}

The circles in Fig.~\ref{fig:combined_all} are the maximum temperature differences for each measurement point for all the replicate cooling events. If water and CAFs have the exact same temperature differences, all of the points would fall along solid black line in the figure. Circles that vary from that line represent measurements where either water (above) or CAFS (below) cooled more signifcantly than the other agent. The gray area surrounding the solid black line, represents measurement uncertainty (\ref{subsec:Uncertainty}). For data that fall within this area, there is an inability to determine which agent was more or less effective than the other. The red dashed line is the line the describes the data by minimizing the differences between the two suppression agents through a linear approximation of the data. Since this line falls within the gray area there is an inability to distinguish, with statistical confidence, a difference between water or CAFs to cool gases by combining all of the replicate tests.

@@

@@

\chapter{Acknowledgments}
\label{chap:acknowledgments}

The authors would like to thank Roy McLane (currently with Thermal Fabrication), Kristopher Overholt (currently with Continuum Analytics), Keith Stakes (currently with UL FSRI), Adam Barowy (currently with UL), Kelly Opert (currently with UL) and Jay McElroy (retired) from NIST for their contributions in preparing and conducting these experiments. These experiments could not have been conducted without them.

The Delaware County Emergency Services Training Center provided the location and logistical support for these experiments. The Delaware County team was under the direction of Kerby Kerber.

Thank you for assisting NIST in generating data that will assist the fire service in making tactical choices.
